Quantum H theorem, variational entropy, and relaxation processes

### 2020-12-21 tex/quantum.tex
Rewording: up tp sec. 2.3.1

The $H$ functional in quantum mechanics presented by Tolman \cite{bib:tolman}
is defined as
%
\begin{equation}\label{quantumh}
    H=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, 
\end{equation}
%
where $n_i$ is the occupation number and $g_i$ is an energy cell. Also, an
equation for the rate of change in the number of particles in group $\kappa$ is
%
\begin{eqnarray}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}),\label{changen}
\end{eqnarray}
%
where $A_{\kappa\lambda,\mu\nu}$ is a tensor expressed as
%
\begin{equation}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon},
\end{equation}
%
where $\epsilon$ is the energy and $|I_1-I_2|^2=|V_{mn,kl}|^2$ where
$V_{mn,kl}$ is the perturbation matrix of binary particles. Those expressions
are obtained by the perturbation theory.

\textcolor{red}{Tolman's proposal for the quantum $H$-theorem establishes that
if we have a quantum gas that satisfies the rate of change equation
(\ref{changen})\footnote{It is important to remark that the equal a priori
probabilities and random a priori phases hypothesis are implicit in the rate of
change equation (\ref{changen}) and the random a priori phases hypothesis is
considered as the quantum analogous to the molecular chaos hypothesis
\cite{bib:das2018}}, then the $H$ functional (\ref{quantumh}) satisfies that
%
\begin{equation}
    \frac{dH}{dt}\leq 0,
\end{equation}
%
and as a consequence of this theorem, when the system is in equilibrium,
$\frac{dH}{dt}=0$.\\
The proof of this theorem is in the same way as in the classical version. Also,
when the system is in equilibrium, the distribution functions that correspond
to the equilibrium state are the Fermi-Dirac and the Bose-Einstein
distribution. Those distributions can be obtained from a sufficient condition
for the equilibrium state
%
\begin{equation}
    \ln \frac{n_{\kappa}}{g_{\kappa}\pm n_{\kappa}}+\ln \frac{n_{\lambda}}{g_{\lambda}\pm n_{\lambda}}=\ln \frac{n_{\mu}}{g_{\mu}\pm n_{\nu}}+\ln \frac{n_{\nu}}{g_{\nu}\pm n_{\nu}},
\end{equation}
%
where the upper sign refers to the Bose-Einstein distribution and the lower
sign refers to the Fermi-Dirac distribution.}

In addition to this, Tolman showed that (\ref{quantumh}) can be reduced in the
limit of high-energy levels to
%
\begin{equation}
    H = \sum_{\kappa} (n_{\kappa} \ln n_{\kappa} - n_{\kappa} \ln g_{\kappa}). \label{reduce-h}
\end{equation}
%
This expression also can be obtained also from the Boltzmann $H$ functional through defining
%
\begin{equation}
    f=\frac{n_{\kappa}}{ g_{\kappa}}.
\end{equation}
%
%--------------------------

@@

Maximizing the variational entropy $\Ss$ using the Lagrange multipliers method,
we obtain
%
\begin{eqnarray}
&&\ln \left(\frac{1\mp f_{Mn}(\epsilon_{n},t)}{f_{Mn}(\epsilon_{n},t)} \right)=-\alpha_M(t)-\beta_M(t) \epsilon_{n}, \label{relation}\\ &&\Rightarrow \ \ f_{Mn}(\epsilon_{n},t)=\frac{1}{e^{-\alpha_M(t)-\beta_M(t) \epsilon_{n}}\pm 1} \equiv \bar{f}_{Mn}(\epsilon_{n},t) \label{distributionequilibrium}.
\end{eqnarray}
%
where $\alpha_M(t), \beta_M(t)$ are the Lagrange multipliers. We can obtain
from the variational procedure that
%
\begin{equation}
    \alpha_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{N}_M}, \ \ \ \beta_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{E}_M},\label{multipliers}
\end{equation}
%
where we used that the $H$ functional in equilibrium is proportional to the
entropy of the system.
In other hand, in the expression (\ref{multipliers}), we expect that the
variation of the logarithm with the number of particles by cells is indifferent
from a particular cell. With this, we conclude 
%
\begin{equation}
    \alpha_M=\alpha; \ \ \ \ \beta_M=\beta.
\end{equation}
%
In equilibrium, the energy and the Lagrange multiplier do not depend on the
cell number and time, therefore
%
\begin{equation}
    \bar f_{Mn}(\epsilon_{n},t)=\bar f_n(\epsilon_{n}) =\frac{1}{e^{-\alpha-\beta \epsilon_n}\pm 1}.
\end{equation}
%
If one compares those multipliers with the mean energy and the mean particle
number from the statistical mechanic's results, we obtain
%
\begin{equation}
    \alpha=\frac{\mu}{kT}\equiv \bar{\alpha}; \ \ \ \ \beta=-\frac{1}{kT}\equiv \bar{\beta},
\end{equation}
%
and finally
%
\begin{eqnarray}
    \bar{f}_{n}(\epsilon_{n})&=&\frac{1}{e^{(\frac{{\epsilon_n}-\bar{\mu}}{kT})}\pm 1}\equiv \bar{f}_{n}.
\end{eqnarray}
%
We proved, when the system is in equilibrium, the Lagrange multipliers don't
depend on the cell number. In other words, Lagrange multipliers are homogeneous
in the system.

>>>>

The $H$-functional in quantum mechanics presented by Tolman \cite{bib:tolman}
is defined as
%
\begin{equation}\label{quantumh}
    H=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, 
\end{equation}
%
where $n_i$ is the occupation number and $g_i$ is an energy cell. Also, an
equation for the rate of change in the number of particles in group $\kappa$ is
%
\begin{eqnarray}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}),\label{changen}
\end{eqnarray}
%
where $A_{\kappa\lambda,\mu\nu}$ is a tensor expressed as
%
\begin{equation}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon},
\end{equation}
%
where $\epsilon$ is the energy and $|I_1-I_2|^2=|V_{mn,kl}|^2$ where
$V_{mn,kl}$ is the perturbation matrix of binary particles. Those expressions
are obtained by the perturbation theory.

\textcolor{red}{Tolman's proposal for the quantum $H$-theorem establishes that
if we have a quantum gas that satisfies the rate of change equation
(\ref{changen})\footnote{It is important to remark that the equal a priori
probabilities and random a priori phases hypothesis are implicit in the rate of
change equation (\ref{changen}) and the random a priori phases hypothesis is
considered as the quantum analogous to the molecular chaos hypothesis
\cite{bib:das2018}}, then the $H$-functional (\ref{quantumh}) satisfies that
%
\begin{equation}
    \frac{dH}{dt}\leq 0,
\end{equation}
%
and as a consequence of this theorem, when the system is in equilibrium,
$\frac{dH}{dt}=0$.\\
The proof of this theorem is in the same way as in the classical version. Also,
when the system is in equilibrium, the distribution functions that correspond
to the equilibrium state are the Fermi-Dirac and the Bose-Einstein
distribution. Those distributions can be obtained from a sufficient condition
for the equilibrium state
%
\begin{equation}
    \ln \frac{n_{\kappa}}{g_{\kappa}\pm n_{\kappa}}+\ln \frac{n_{\lambda}}{g_{\lambda}\pm n_{\lambda}}=\ln \frac{n_{\mu}}{g_{\mu}\pm n_{\nu}}+\ln \frac{n_{\nu}}{g_{\nu}\pm n_{\nu}},
\end{equation}
%
where the upper sign refers to the Bose-Einstein distribution and the lower
sign refers to the Fermi-Dirac distribution.}

In addition to this, Tolman showed that (\ref{quantumh}) can be reduced in the
limit of high-energy levels to
%
\begin{equation}
    H = \sum_{\kappa} (n_{\kappa} \ln n_{\kappa} - n_{\kappa} \ln g_{\kappa}). \label{reduce-h}
\end{equation}
%
This expression also can be obtained also from the Boltzmann $H$-functional through defining
%
\begin{equation}
    f=\frac{n_{\kappa}}{ g_{\kappa}}.
\end{equation}
%
%--------------------------

@@

Maximizing the variational entropy $\Ss$ using the Lagrange multipliers method,
we obtain
%
\begin{eqnarray}
&&\ln \left(\frac{1\mp f_{Mn}(\epsilon_{n},t)}{f_{Mn}(\epsilon_{n},t)} \right)=-\alpha_M(t)-\beta_M(t) \epsilon_{n}, \label{relation}\\ &&\Rightarrow \ \ f_{Mn}(\epsilon_{n},t)=\frac{1}{e^{-\alpha_M(t)-\beta_M(t) \epsilon_{n}}\pm 1} \equiv \bar{f}_{Mn}(\epsilon_{n},t) \label{distributionequilibrium}.
\end{eqnarray}
%
where $\alpha_M(t), \beta_M(t)$ are the Lagrange multipliers. We can obtain
from the variational procedure that
%
\begin{equation}
    \alpha_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{N}_M}, \ \ \ \beta_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{E}_M},\label{multipliers}
\end{equation}
%
where we used that the $H$-functional in equilibrium is proportional to the
entropy of the system.
In other hand, in the expression (\ref{multipliers}), we expect that the
variation of the logarithm with the number of particles by cells is indifferent
from a particular cell. With this, we conclude 
%
\begin{equation}
    \alpha_M=\alpha; \ \ \ \ \beta_M=\beta.
\end{equation}
%
In equilibrium, the energy and the Lagrange multiplier do not depend on the
cell number and time, therefore
%
\begin{equation}
    \bar f_{Mn}(\epsilon_{n},t)=\bar f_n(\epsilon_{n}) =\frac{1}{e^{-\alpha-\beta \epsilon_n}\pm 1}.
\end{equation}
%
If one compares those multipliers with the mean energy and the mean particle
number from the statistical mechanic's results, we obtain
%
\begin{equation}
    \alpha=\frac{\mu}{kT}\equiv \bar{\alpha}; \ \ \ \ \beta=-\frac{1}{kT}\equiv \bar{\beta},
\end{equation}
%
and finally
%
\begin{eqnarray}
    \bar{f}_{n}(\epsilon_{n})&=&\frac{1}{e^{(\frac{{\epsilon_n}-\bar{\mu}}{kT})}\pm 1}\equiv \bar{f}_{n}.
\end{eqnarray}
%
We proved, when the system is in equilibrium, the Lagrange multipliers don't
depend on the cell number. In other words, Lagrange multipliers are homogeneous
in the system.
